---
title: "Biodiversity in Vancouver Parks"
author: "Lesley Miller"
date: "07/06/2020"
output: 
  html_document:
    toc: true
    theme: darkly
    pandoc_args: --webtex
always_allow_html: true
bibliography: citations.bib
runtime: shiny
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = here::here())
library(shiny)
suppressPackageStartupMessages(library(tidyverse))
library(sf)
library(knitr)
library(gridExtra)
library(grid)
library(shinythemes)
library(hrbrthemes)

```

```{r load_data}
# load gbif
gbif <- st_read('data/clean_data/vancouver_gbif.shp', quiet = TRUE)

# load all parks w/metrics 
parks <- read_csv("data/clean_data/all_parks_with_metrics.csv", col_types = cols()) # 


# load clipped_gbif
clipped_gbif <- st_read('data/clean_data/clipped_gbif.shp', quiet = TRUE)
clipped_gbif <- clipped_gbif %>% 
      rename(basis_of_record = basis_of_r)

# load outlier parks
outlier_parks <-  st_read('data/clean_data/park_outliers.shp', quiet = TRUE)
outlier_parks <- outlier_parks %>% 
      rename(park_primary_use = park_prima,
             species_counts = species_co,
             species_richness = species_ri )

# load parks with no outliers 
parks_with_no_outliers <- st_read('data/clean_data/parks_with_no_outliers.shp', quiet = TRUE)
parks_with_no_outliers <- parks_with_no_outliers %>% 
      rename(park_primary_use = park_prima,
             species_counts = species_co,
             species_richness = species_ri)
```


## Introduction
The primary interest of this project was to explore the intersection between Vancouver, BC parks and geotagged species occurrence data obtained from the [Global Biodiversity Information Facility](https://www.gbif.org/), known from here on as GBIF. Any Vancouverite would be proud of our city parks and greenspaces as they serve as immense community resources with multifunctional benefits. Given that urban parks may be the primary interface with nature and wildlife for a great number of residents, this project attempted to answer how much species occurrence data has been collected in Vancouver parks over the last 10 years. The analysis answers 3 main questions. 

1) What proportion of Vancouver's biodiversity data is recorded in an urban park?
2) Which parks contain the most observations?
3) Which parks contain the largest number of different species?

To answer these questions, the analysis makes use of two main data sources: Species occurrence data from [GBIF](https://www.gbif.org/) and spatial parks data from the [BC Data Catalogue](https://catalogue.data.gov.bc.ca/dataset/local-and-regional-greenspaces). The interactive tables below allows a peak at the data underlying the analysis. Only the first few rows of each dataset are shown. 

The GBIF contained 5,110 different observations and 882 unique species spanning 11 years from 2009 to 2019. All records are from human observation. As can be seen in the interactive GBIF table, each record contains the full taxonomy, geolocation, as well as the date and time an observation was recorded. 

The parks data conains information for 234 parks having a range of primary park uses; the main use is just **Park** but the dataset also contains other types of greenspaces like athletic parks, golf courses and playgrounds. 

```{r park_use}
park_use <- as.data.frame(table(parks$park_prima))
colnames(park_use) <- c("Park Type", "Frequency")
kable(park_use, caption = "Primary Uses of Vancouver Parks")
```



## Data Sources {.tabset .tabset-pills}

### GBIF Data
```{r gbif_data}
# gbif data table

shinyApp(
  ui = fluidPage(
    theme = shinytheme("cerulean"),
    fluidRow(
      column(12,
        dataTableOutput('table1')
      )
    )
  ),
  server = function(input, output) {
    output$table1 <- renderDataTable(head(clipped_gbif,50))
  }
)

```

### Parks Data
```{r parks_data}

# DataTables 
shinyApp(
  ui = fluidPage(
    theme = shinytheme("cerulean"),
    fluidRow(
      column(12,
        dataTableOutput('table2')
      )
    )
  ),
  server = function(input, output) {
    output$table2 <- renderDataTable(parks_with_no_outliers)
  }
)

```


# Data Processing Pipeline 

## Data Cleaning 
The raw GBIF data was first obtained through their API and then clipped to a gridded boundary obtained from the [City of Vancouver Open Data Portal]('https://opendata.vancouver.ca/explore/dataset/facet-grid-boundaries/map/?location=12,49.2562,-123.1389'). The data was then cleaned by removal of unnecessary columns. The parks dataset underwent only minimal processing to retain only necessary columns for downstream analysis. 

## Data Preparation 
Since the analysis is visual, the data was then processed to produce new data that could be fed into a multi-layer map. The GBIF data points were clipped to within the park boundaries so that total observation counts as well as the count of unique species per park could be calculated. The clipped GBIF data and the parks along with its species metrics then became data layers in a **keplergl** map available [here for exploration](https://aromatic-toast.github.io/vancouver_park_biodiversity/results/final_report_map.html).

## Data Visualization 
The primary visualization product is a spatial map of Vancouver that includes park polygons and species observation data that was processed by the data pipeline. The map was produced using [@keplergl] and [can be found here](https://aromatic-toast.github.io/vancouver_park_biodiversity/results/final_report_map.html). A detailed walk-through of ways to explore a **keplergl** map is detailed [here](https://github.com/aromatic-toast/vancouver_park_biodiversity/blob/master/README.md). The map allows a user to zoom in on areas of interest as well as toggle data layers on/off using the hidden tool bar on the left hand side of the map. A tooltip displays the count metrics for parks (e.g observation count and unique species count) while the tool tip for a species observation includes the full taxonomy information as well as the type of record that was recorded. 

This analysis was carried out with the amazing help from the following open source **R** packages: **R** [@R], **tidyverse** [@tidyverse], **here** [@here], **gridExtra** [@gridExtra] and **sf** [@sf]. 

And the following **Python** packages: **Python** [@Python], **GeoPandas** [@geopandas0.7.0], **Pandas** [@pandas] and **keplergl** [@keplergl].

# Results 

##  Observation Hotspots {.tabset .tabset-pills}

***What proportion of Vancouver's biodiversity data is recorded in an urban park?***

The pie chart in **Proportion in Parks** indicates that about half of the GBIF data for Vancouver was observed in a park with the true percentage about 51%. 

***Which parks contain the most observations?***

Unsurprisingly, **Stanley Park** captured nearly a third (~ 28%) of the 5,110 total observations and **Pacific Spirit** captured about a fifth (22.8%). The observation counts for these two parks can be viewed under the **Outliers** tab. Together they constitute about half of the observations. The other half of the observations are distributed across 69 other parks; 165 parks contained no species observations at all. The **No Outliers** shows the observation counts for the top 10 parks after excluding the two outliers. When not considering the outliers, **McCleery Golf Course** comes in first place for highest observations followed by **Jericho Beach Park** and **Queen Elizabeth Park**. 

### Proportion In Parks
```{r}
# get relative proportions of observations inside and outside parks 
observed_in_park <- nrow(clipped_gbif)
outside_park <-  nrow(gbif) - observed_in_park

species_in_out_df <- tibble(count_observed_in_park = c('yes', 'no'),
                            count = c(observed_in_park, outside_park))

pie <- species_in_out_df %>% 
  ggplot(aes(x = factor(1), fill = count_observed_in_park)) +
  geom_bar(width = 1) +
  coord_polar("y") + 
  theme_light() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(face = "bold")) + 
  labs(title = "Relative Proportion of Observations in Parks") +
  scale_fill_manual(values = c("lightgreen", "darkgreen"))

pie
```

### Outliers
```{r}
#### outliers #### 
plot2_1 <- outlier_parks %>% 
      ggplot(aes(x = park_name, species_counts)) +
      geom_col() + 
      labs(title = "Outlier Observation Hotspots",
           x = 'Park Name',
           y = 'Count of Observations') +
      coord_flip() + 
  theme_ipsum()
plot2_1
```


### No Outliers
```{r}
# Question 2: What are the top 10 parks with the highest observations?
obs_hotspots_no_outliers <- parks_with_no_outliers %>% 
      arrange(desc(species_counts)) %>% 
      head(10) %>% 
      select(park_name, species_counts, species_richness)

plot2 <- obs_hotspots_no_outliers %>% 
      ggplot(aes(x = reorder(park_name, species_counts), y = species_counts)) + 
      geom_col() + 
      labs(title = "Top Ten Observation Hotspots",
           x = 'Park Name',
           y = 'Count of Observations') +
      coord_flip() +
  theme_ipsum()
plot2
```






## Highest Unique Species {.tabset .tabset-pills}


***Which parks contain the largest number of different species?***

The distribution of unique species is very similar to the top ten observation counts. The two winning outliers are again **Stanley Park** and **Pacific Spirit**. For the top ten excluding outliers, **McCleery Golf Course** wins again with the highest number of different species observed. After the winner, the rankings shift a bit compared to the observation counts with nine parks being in the the top ten for both metrics. However, for **Point Grey Golf and Country Club**, it makes the top ten for higher count of unique species but does not appear in the top ten for total observations. 

### Outliers
```{r}
### outliers species richnesss ###
plot3_1 <- outlier_parks %>% 
      ggplot(aes(x = reorder(park_name, species_richness), y = species_richness)) + 
      geom_col() + 
      labs(title = "Outlier Parks for Species Richness",
           x = 'Park Name',
           y = 'Unique Species Count') +
      coord_flip() + 
  theme_ipsum()
plot3_1
```


### No Outliers
```{r}
##### Question 3: which parks have the most species richness? ##### 
top_species_rich <- parks_with_no_outliers %>% 
      arrange(desc(species_richness)) %>% 
      head(10) %>% 
      select(park_name, species_counts, species_richness)

plot3 <- top_species_rich %>% 
      ggplot(aes(x = reorder(park_name, species_richness), y = species_richness)) + 
      geom_col() + 
      labs(title = "Top Ten Parks for Species Richness",
           x = 'Park Name',
           y = 'Unique Species Count') +
      coord_flip() +
  theme_ipsum()
  
plot3
```


# Conclusion
This project explored the urban parks and species observations of Vancouver, BC. A key curiosity at the start of this project was how much citizen science data is recorded in Vancouver parks. Given how sparse the GBIF data is for Vancouver, there is only ~5,000 observations over an 11 year period, it was thought at the beginning that perhaps not many observations would be found in these places or perhaps they would be extremely clustered in only a few parks. It turns out that just over half of the data collected was found within parks, an unexpected result. This reinforces the presumption at the start of this project that urban parks serve as a gateway for residents to observe and interact with their natural environment. Though half the observations are found in only two parks, there are a number of other parks scattered across the landscape that serve as observation hotspots. 

A future analysis would look at not only raw observation counts per park but also incorporate park size as most of the large parks also tend ot have large numbers of observations. An additional metric might compare parks based on their observations per unit area. 

This brief dive into Vancouver parks underscores the incredible asset of **Stanley Park** to our urban landscape and provides support for the continued stewardship of these most precious resources. 

# References
